---
title: "NFL Data"
output:
  pdf_document: default
  html_document: default
date: "2025-12-15"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(nflfastR)
```

### Executive Summary

A persistent inefficiency in NFL roster construction and free-agent evaluation is the overvaluation of raw rookie volume. Traditional scouting heavily factors in Year 1 counting stats without isolating the underlying, repeatable skills that actually drive future success. Furthermore, evaluations often fail to account for the systemic bias of draft capital, where highly drafted players are force-fed opportunities regardless of their baseline efficiency.

This project utilizes play-by-play data to build a multiple linear regression model that predicts Year 2 Wide Receiver production. By isolating First Down Rate and adjusting for Draft Capital, this model successfully explains 41.4% of the variance in Year 2 receiving yards, providing a mathematical framework for identifying undervalued trade targets and optimizing salary cap allocations.

### Methodology & Data Pipeline

-   **Data Ingestion:** Utilized nflreadr to scrape and aggregate 11 seasons (2012–2023) of play-by-play data and official NFL roster files.

-   **Feature Engineering:** Aggregated weekly logs into seasonal totals to calculate true rookie-year volume and a custom efficiency metric: First Down Rate, defined as receiving first downs per target.

-   **Contextual Variables:** Merged roster data to assign each player's exact draft pick number, explicitly modeling the front-office "sunk-cost fallacy" to see how draft pedigree influences Year 2 opportunities.

-   **Modeling:** Deployed a multiple linear regression model predicting Year 2 Receiving Yards based on Year 1 Targets, Year 1 First Down Rate, and Draft Pick Number.

### Statistical Findings

The model yielded a highly actionable output, parsing exactly which traits translate year-over-year.

-   **Overall Predictive Power**: Despite the extreme volatility of the NFL (injuries, coaching changes, quarterback turnover), this model successfully explains over 41% of a player's Year 2 yardage utilizing strictly three rookie-year variables.

-   **First Down Rate:** The data definitively proves that the ability to move the chains is a highly significant, repeatable skill. The regression coefficient (861.27) dictates that if two rookies see the exact same target volume, the player who converts targets into first downs at a 10% higher rate projects for an additional 86 yards in Year 2.

-   **The Sunk-Cost Bias - Draft Capital :** The model returned a coefficient of -0.3277 for draft pick number. This mathematically captures the systemic bias in NFL coaching: the lower a player's draft pick number (e.g., Pick 1 vs. Pick 100), the higher their projected Year 2 yardage, completely independent of their actual rookie-year efficiency. Every 100 picks a player falls in the draft inherently costs them roughly 33 expected yards in their second season.

-   **Volume Baseline:**

    Earning the trust of the coaching staff to get on the field remains the ultimate baseline. For every target earned in Year 1, a player projects for 7.09 yards in Year 2.

```{r}
# 
# PROJECT: YEAR 2 BREAKOUT PREDICTOR 
# 
library(tidyverse)
library(nflreadr)

print("1. Downloading Data...")
weekly_stats <- load_player_stats(seasons = 2012:2023)
print(paste("Rows downloaded:", nrow(weekly_stats)))

print("2. Aggregating to Season Totals...")
# We use 'first(player_display_name)' to ensure name changes don't duplicate rows
season_stats <- weekly_stats %>%
  group_by(player_id, season) %>%
  summarize(
    player_name = first(player_display_name), 
    targets = sum(targets, na.rm = TRUE),
    rec_yards = sum(receiving_yards, na.rm = TRUE),
    rec_epa = sum(receiving_epa, na.rm = TRUE),
    .groups = "drop"
  )
print(paste("Unique player-seasons:", nrow(season_stats)))

print("3. Finding Rookie Years...")
career_starts <- season_stats %>%
  group_by(player_id) %>%
  summarize(
    rookie_year = min(season),
    career_targets = sum(targets, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(rookie_year > 2012, career_targets >= 50)
print(paste("Valid rookies found:", nrow(career_starts)))

print("4. Mapping Year 1 vs Year 2...")
wr_timeline <- season_stats %>%
  inner_join(career_starts, by = "player_id") %>%
  mutate(career_year = season - rookie_year + 1)

year_1 <- wr_timeline %>%
  filter(career_year == 1, targets >= 30) %>%
  select(player_id, player_name, targets_y1 = targets, 
         rec_yards_y1 = rec_yards, rec_epa_y1 = rec_epa) %>%
  mutate(epa_per_target_y1 = rec_epa_y1 / targets_y1)
print(paste("Players with 30+ targets in Year 1:", nrow(year_1)))

year_2 <- wr_timeline %>%
  filter(career_year == 2) %>%
  select(player_id, rec_yards_y2 = rec_yards)
print(paste("Players who also played Year 2:", nrow(year_2)))

print("5. Merging for Model...")
final_model_data <- inner_join(year_1, year_2, by = "player_id") %>%
  drop_na()
print(paste("FINAL PLAYERS FOR MODEL:", nrow(final_model_data)))

# 6. Safety Check & Model Execution
if(nrow(final_model_data) > 0) {
  print("SUCCESS: Running the Regression Model!")
  breakout_model <- lm(rec_yards_y2 ~ epa_per_target_y1 + rec_yards_y1, data = final_model_data)
  print(summary(breakout_model))
  
  # Plot
  p <- ggplot(final_model_data, aes(x = epa_per_target_y1, y = rec_yards_y2)) +
    geom_point(alpha = 0.6, color = "darkblue", size = 2) +
    geom_smooth(method = "lm", color = "red", linetype = "dashed") +
    labs(
      title = "Rookie Efficiency Translating to Year 2 Production",
      subtitle = "Higher EPA per Target in Year 1 dictates Year 2 leaps.",
      x = "Year 1 Efficiency (EPA per Target)",
      y = "Year 2 Total Receiving Yards"
    ) +
    theme_minimal()
  print(p)
} else {
  print("ERROR: 0 players made it to the final dataset. Please copy and paste the print outputs above so I can see exactly where the pipeline failed!")
}
```

```{r}
 
library(tidyverse)
library(nflreadr)

print("1. Downloading Data...")
weekly_stats <- load_player_stats(seasons = 2012:2023)

print("2. Aggregating to Season Totals & Engineering Advanced Metrics...")
season_stats <- weekly_stats %>%
  group_by(player_id, season) %>%
  summarize(
    player_name = first(player_display_name), 
    targets = sum(targets, na.rm = TRUE),
    receptions = sum(receptions, na.rm = TRUE),
    rec_yards = sum(receiving_yards, na.rm = TRUE),
    rec_epa = sum(receiving_epa, na.rm = TRUE),
    # NEW VARIABLES PULLED FROM nflfastR:
    air_yards = sum(receiving_air_yards, na.rm = TRUE),
    yac = sum(receiving_yards_after_catch, na.rm = TRUE),
    first_downs = sum(receiving_first_downs, na.rm = TRUE),
    .groups = "drop"
  )

print("3. Finding Rookie Years...")
career_starts <- season_stats %>%
  group_by(player_id) %>%
  summarize(
    rookie_year = min(season),
    career_targets = sum(targets, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(rookie_year > 2012, career_targets >= 50)

print("4. Mapping Year 1 vs Year 2 Features...")
wr_timeline <- season_stats %>%
  inner_join(career_starts, by = "player_id") %>%
  mutate(career_year = season - rookie_year + 1)

# ENGINEER YEAR 1 FEATURES
year_1 <- wr_timeline %>%
  filter(career_year == 1, targets >= 30) %>%
  mutate(
    epa_per_target_y1 = rec_epa / targets,
    aDOT_y1 = air_yards / targets,
    fd_rate_y1 = first_downs / targets,
    yac_per_rec_y1 = ifelse(receptions > 0, yac / receptions, 0)
  ) %>%
  select(player_id, player_name, targets_y1 = targets, 
         epa_per_target_y1, aDOT_y1, fd_rate_y1, yac_per_rec_y1)

# ENGINEER YEAR 2 TARGET (We are now predicting Year 2 Efficiency!)
year_2 <- wr_timeline %>%
  filter(career_year == 2, targets >= 30) %>% # Require 30 targets in Y2 to remove noisy 1-target data
  mutate(epa_per_target_y2 = rec_epa / targets) %>%
  select(player_id, epa_per_target_y2)

print("5. Merging for Model...")
final_model_data <- inner_join(year_1, year_2, by = "player_id") %>%
  drop_na()

print(paste("FINAL PLAYERS FOR MODEL:", nrow(final_model_data)))

# 6. The "Kitchen Sink" Regression
if(nrow(final_model_data) > 0) {
  print("SUCCESS: Running the Multiple Regression Model!")
  
  # We are testing 4 different Year 1 metrics to see what predicts Year 2 Efficiency
  efficiency_model <- lm(epa_per_target_y2 ~ targets_y1 + aDOT_y1 + fd_rate_y1 + yac_per_rec_y1, 
                         data = final_model_data)
  
  print(summary(efficiency_model))
  
} else {
  print("ERROR: Pipeline failed.")
}
```

```{r}
# 
# ADJUSTED BREAKOUT MODEL
# 
library(tidyverse)
library(nflreadr)

print("1. Downloading Stats and Rosters...")
weekly_stats <- load_player_stats(seasons = 2012:2023)
rosters <- load_rosters(seasons = 2012:2023) %>%
  # Grab their draft pick number
  select(player_id = gsis_id, draft_number) %>%
  distinct(player_id, .keep_all = TRUE) %>%
  # If they were undrafted, we assign them pick '300' (past the 7th round)
  mutate(draft_number = replace_na(as.numeric(draft_number), 300))

print("2. Aggregating to Season Totals...")
season_stats <- weekly_stats %>%
  group_by(player_id, season) %>%
  summarize(
    player_name = first(player_display_name), 
    targets = sum(targets, na.rm = TRUE),
    rec_yards = sum(receiving_yards, na.rm = TRUE),
    first_downs = sum(receiving_first_downs, na.rm = TRUE),
    .groups = "drop"
  )

print("3. Finding Rookie Years...")
career_starts <- season_stats %>%
  group_by(player_id) %>%
  summarize(
    rookie_year = min(season),
    career_targets = sum(targets, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(rookie_year > 2012, career_targets >= 50)

print("4. Mapping Features & Draft Capital...")
wr_timeline <- season_stats %>%
  inner_join(career_starts, by = "player_id") %>%
  mutate(career_year = season - rookie_year + 1) %>%
  # Merge in the Draft Pick data
  inner_join(rosters, by = "player_id")

# ENGINEER YEAR 1 FEATURES
year_1 <- wr_timeline %>%
  filter(career_year == 1, targets >= 30) %>%
  mutate(fd_rate_y1 = first_downs / targets) %>%
  select(player_id, player_name, targets_y1 = targets, 
         fd_rate_y1, draft_number)

# ENGINEER YEAR 2 TARGET (Predicting Total Year 2 Yards)
year_2 <- wr_timeline %>%
  filter(career_year == 2) %>%
  select(player_id, rec_yards_y2 = rec_yards)

print("5. Merging for Model...")
final_model_data <- inner_join(year_1, year_2, by = "player_id") %>%
  drop_na()

print(paste("FINAL PLAYERS FOR MODEL:", nrow(final_model_data)))

# 6. The Context-Adjusted Regression
if(nrow(final_model_data) > 0) {
  print("SUCCESS: Running the Ultimate Prediction Model!")
  
  # Predicting Year 2 Yards based on Volume, Skill, and Draft Pedigree
  ultimate_model <- lm(rec_yards_y2 ~ targets_y1 + fd_rate_y1 + draft_number, data = final_model_data)
  
  print(summary(ultimate_model))
  
} else {
  print("ERROR: Pipeline failed.")
}
```
